---
title: "Emergent constraint on beta-GPP"
author: "Beni Stocker"
date: "2023-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


## Read data

Data obtained by Trevor by email of 26 January 2022 (attachment `modelTerms.zip`). Unpacked and added to this repo.
```{r}
df_beta <- read_csv("../data/T_betaGPP.csv") |> 
  pivot_longer(everything(), values_to = "beta", names_to = "model")

df_sland <- read_csv("../data/T_Sland.csv") |> 
  pivot_longer(everything(), values_to = "sland", names_to = "model")

df <- df_beta |> 
  left_join(df_sland, by = "model") |> 
  
  # remove row with missing value (for model VEGAS)
  drop_na()
```

## Regression

Define prediction error function based on methods in Cox et al., 2013.

Add line for prediction error $\sigma_f(x)$ manually.
```{r}
# sum of square errors
linmod <- lm(beta ~ sland, data = df)

# prediction error (sigma_f)
get_prediction_error <- function(x, x_mean, x_var, linmod){
  sse <- sum(linmod$residuals^2)
  s_squared <- (1 / (length(linmod$residuals) - 2)) * sse
  out <- sqrt(s_squared) * sqrt(1 + 1/length(sse) + (x - x_mean)^2 / (length(sse) * x_var))
  return(out)
}

# prediction +/- prediction error
predict_range_upper <- function(x, linmod){
  predict(linmod, data.frame(sland = x)) + 
    get_prediction_error(x, 
                         x_mean = mean(x),
                         x_var = var(x),
                         linmod)
}
predict_range_lower <- function(x, linmod){
  predict(linmod, data.frame(sland = x)) -
    get_prediction_error(x, 
                         x_mean = mean(x),
                         x_var = var(x),
                         linmod)
}

df |> 
  ggplot(aes(sland, beta)) + 
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  geom_function(fun = function(x) predict_range_upper(x, linmod = linmod)) +
  geom_function(fun = function(x) predict_range_lower(x, linmod = linmod)) +
  theme_classic()
```


## Unconstrained distribution

The unconstrained probability distribution of $y$ is given by a normal distribution with mean and standard deviations given by the vector of $\beta$ of the models.

```{r}
df |> 
  ggplot() +
  geom_histogram(aes(beta, ..density..), bins = 5, color = "black", fill = "grey70") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(df$beta, na.rm = TRUE), 
                            sd = sd(df$beta, na.rm = TRUE)),
                color = "red") +
  theme_classic()
```

## Distribution of S_land

The land sink is assumed to be normally distributed with mean = X and SD = Y. For demo, using mean = 3 and SD = 0.1.

@Trevor: Could you send me the updated data with cumulative Sland values from models and report the  mean and standard deviation of the Sland used for the constraint?

```{r}
prob_x <- function(x){
  dnorm(x, mean = 3.0, sd = 0.1)
}

ggplot() +
  geom_function( fun = prob_x) + 
  xlim(0, 5)
```

## Constrained probability distribution

$$
P(y) = \int_{-\infty }^{+\infty} P(y|x) P(x)
$$
```{r}
# P(y|x) is a Gaussian function with mean = prediction and variance = prediction error
prob_y_given_x <- function(x, y, x_mean, x_var, linmod){
  
  # out <- dnorm(y, 
  #              mean = predict(linmod, data.frame(sland = x)),
  #              sd = get_prediction_error(x, x_mean, x_var, linmod))
  
  # or "manually":
  f_x <- predict(linmod, data.frame(sland = x))
  sigma_f <- get_prediction_error(x, x_mean, x_var, linmod)
  out <- (1 / sqrt(2 * pi * sigma_f^2)) * exp(-(y - f_x)^2 / (2 * sigma_f^2))  # error in Cox formula: missing square
  
  return(out)
}

d_constrained_prob <- function(x, y, x_mean, x_var, linmod){
  prob_y_given_x(x, y, x_mean, x_var, linmod) * prob_x(x)
}

constrained_prob <- function(y, x_mean, x_var, linmod){
  integrate(d_constrained_prob, 
            lower = -10, 
            upper = 10, 
            y = y, 
            x_mean = x_mean, 
            x_var = x_var, 
            linmod = linmod
            )$value
}
```

Plot unconstrained (black) and constrained (red) probability distributions.
```{r}
# create values for constrained probability
df_prob <- tibble(y = seq(from = -1, to = 2, by = 0.025)) |> 
  rowwise() |> 
  mutate(prob_y = constrained_prob(y, 
                                   x_mean = mean(df$sland), 
                                   x_var = var(df$sland), 
                                   linmod = linmod))

df |> 
  ggplot() +
  geom_histogram(aes(beta, ..density..), 
                 bins = 30, 
                 color = "black", 
                 fill = "grey70") +
  stat_function(fun = dnorm, 
                args = list(mean = mean(df$beta, na.rm = TRUE), 
                            sd = sd(df$beta, na.rm = TRUE)),
                color = "black") +
  geom_line(data = df_prob, aes(y, prob_y), color = "red") +
  theme_classic()
```
